---
title: "BreastCancer Wisconsin Diagnostic dataset"
author: "Shravan Kuchkula"
date: "10/22/2017"
output:
  github_document:
    toc: yes
  html_document:
    keep_md: yes
    theme: cosmo
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
---

## Introduction
Features are computed from a digitized image of a fine needle	aspirate (FNA) of a breast mass.  They describe	characteristics of the cell nuclei present in the image.

## Data set
Breast Cancer Wisconsin data set from the [*UCI Machine learning repo*](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29) is used to conduct the analysis. 

## Importing and Cleaning the data

Before importing, let's first load the required libraries.

```{r message=FALSE, warning=FALSE}
source('libraries.R')
```

Using read.csv we can download the dataset as shown:

```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data"

# use read_csv to the read into a dataframe
# columnNames are missing in the above link, so we need to give them manually.
columnNames <- c("id","diagnosis","radius_mean","texture_mean","perimeter_mean",
                 "area_mean","smoothness_mean","compactness_mean","concavity_mean",
                 "concave_points_mean","symmetry_mean","fractal_dimension_mean",
                 "radius_se","texture_se","perimeter_se","area_se","smoothness_se",
                 "compactness_se","concavity_se","concave_points_se","symmetry_se",
                 "fractal_dimension_se","radius_worst","texture_worst","perimeter_worst",
                 "area_worst","smoothness_worst","compactness_worst","concavity_worst",
                 "concave_points_worst","symmetry_worst","fractal_dimension_worst")
#wdbc <- read_csv(url, col_names = columnNames, col_types = NULL)
wdbc <- read.csv(url, header = FALSE, col.names = columnNames)
```

Let's take a peak
```{r}
glimpse(wdbc)
```

## Exploratory Data Analysis
Our response variable is diagnosis: Benign (B) or Malignant (M).
We have 3 sets of 10 numeric variables: mean, se, worst

Let's first collect all the 30 numeric variables into a matrix

```{r}
# Convert the features of the data: wdbc.data
wdbc.data <- as.matrix(wdbc[,c(3:32)])

# Set the row names of wdbc.data
row.names(wdbc.data) <- wdbc$id

# Create diagnosis vector
diagnosis <- as.numeric(wdbc$diagnosis == "M")
```

Let's answer some basic questions:

### How many observations are in this dataset ?
```{r}
nrow(wdbc.data)
```

### How many variables/features in the data are suffixed with _mean, _se, _worst?
```{r}
sum(endsWith(colnames(wdbc.data), "_mean"))
sum(endsWith(colnames(wdbc.data), "_se"))
sum(endsWith(colnames(wdbc.data), "_worst"))
```

### How many observations have benign or malignant diagnosis ?
```{r}
table(wdbc$diagnosis)
```
 
### What is the mean of each of the numeric columns ?
```{r}
colMeans(wdbc.data)
```

### What is the sd of each of the numeric columns ?
```{r}
apply(wdbc.data, 2, sd)
```

## PCA
Due to the number of variables in the model, we can try using a dimentionality reduction technique to unveil any patterns in the data. It also helps in visualizing a multi-dimentional dataset like this.

The first step in doing a PCA, is to ask ourselves whether or not the data should be scaled to unit variance. That is, to bring all the numeric variables to the same scale.

Based on the output from `mean` and `sd`, it does appear that some variables have larger variance, e.g. symmetry_se. Hence we should use scaling. (will get this later and prove why we need to scale)

Running PCA:

```{r}
wdbc.pr <- prcomp(wdbc.data, scale = TRUE, center = TRUE)
summary(wdbc.pr)
```

84.73% of the variation is explained by the first five PC's.

### Bi-Plot

Let's create a bi-plot to visualize this:
```{r}
biplot(wdbc.pr)
```

From the above bi-plot of PC1 vs PC2, we can see that all these variables are trending in the same direction and most of them are highly correlated (More on this .. we can visualize this in a corrplot)

Create a scatter plot of observations by components 1 and 2
```{r}
# Scatter plot observations by components 1 and 2
plot(wdbc.pr$x[, c(1, 2)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC2")
```

There is a clear seperation of diagnosis (M or B) that is evident in the PC1 vs PC2 plot.

Let's also take PC1 vs PC3 plot:
```{r}
# Repeat for components 1 and 3
plot(wdbc.pr$x[, c(1,3)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC3")
```

Because principal component 2 explains more variance in the original data than principal component 3, you can see that the first plot has a cleaner cut separating the two subgroups.

### Scree plots

Scree plots can be useful in deciding how many PC's we should keep in the model. Let's create the scree-plots in R. As there is no R function to create a scree-plot, we need to prepare the data for the plot.

```{r}
# Set up 1 x 2 plotting grid
par(mfrow = c(1, 2))

# Calculate variability of each component
pr.var <- wdbc.pr$sdev ^ 2

# Variance explained by each principal component: pve
pve <- pr.var/sum(pr.var)
```

Create a plot of variance explained for each principal component.

```{r}
# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "b")

# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "b")
```

Scree-plots suggest that 80% of the variation in the numeric data is captured in the first 5 PCs.

## LDA
As found in the PCA analysis, we can keep 5 PCs in the model. Our next task is to use the first 5 PCs to build a Linear discriminant function using the `lda()` function in R. 

From the `wdbc.pr` object, we need to extract the first five PC's. To do this, let's first check the variables available for this object.

```{r}
ls(wdbc.pr)
```

We are interested in the `rotation` (also called loadings) of the first five principal components multiplied by the scaled data, which are called `scores` (basically PC transformed data)

```{r}
wdbc.pcs <- wdbc.pr$x[,1:5]
head(wdbc.pcs, 20)
```

Here, the rownames help us see how the PC transformed data looks like.

Now, we need to append the `diagnosis` column to this PC transformed data frame `wdbc.pcs`. Let's call the new data frame as `wdbc.pcst`. 

```{r}
wdbc.pcst <- wdbc.pcs
wdbc.pcst <- cbind(wdbc.pcs, diagnosis)
head(wdbc.pcst)
```

Here, diagnosis == 1 represents malignant and diagnosis == 0 represents benign.


## Splitting the dataset into training/test data
When creating the LDA model, we can split the data into training and test data. Using the training data we can build the LDA function. Next, we use the test data to make predictions. 

```{r}
# Calculate N
N <- nrow(wdbc.pcst)

# Create a random number vector
rvec <- runif(N)

# Select rows from the dataframe
wdbc.pcst.train <- wdbc.pcst[rvec < 0.75,]
wdbc.pcst.test <- wdbc.pcst[rvec >= 0.75,]

# Check the number of observations
nrow(wdbc.pcst.train)
nrow(wdbc.pcst.test)
```

So, 433 observations are in training dataset and 136 observations are in the test dataset. 
We will use the training dataset to calculate the `linear discriminant function` by passing it to the `lda()` function of the `MASS` package. 

```{r message=FALSE}
library(MASS)

wdbc.pcst.train.df <- wdbc.pcst.train

# convert matrix to a dataframe
wdbc.pcst.train.df <- as.data.frame(wdbc.pcst.train)

# Perform LDA on diagnosis
wdbc.lda <- lda(diagnosis ~ PC1 + PC2 + PC3 + PC4 + PC5, data = wdbc.pcst.train.df)
```

Let's summarize the LDA output:

```{r}
wdbc.lda
```

Let's use this to predict by passing the predict function's newdata as the testing dataset.

```{r}
wdbc.pcst.test.df <- wdbc.pcst.test

# convert matrix to a dataframe
wdbc.pcst.test.df <- as.data.frame(wdbc.pcst.test)

wdbc.lda.predict <- predict(wdbc.lda, newdata = wdbc.pcst.test.df)

```

Let's check what functions we can invoke on this predict object:
```{r}
ls(wdbc.lda.predict)
```

Our predictions are contained in the `class` attribute.

```{r}
# print the predictions
(wdbc.lda.predict.class <- wdbc.lda.predict$class)
```

Next, compare the accuracy of these predictions with the original data.

## Cross validation
A simple way to validate the accuracy of our model in predicting diagnosis (M or B) is to compare the test data result to the observed data. Find the proportion of the errors in prediction and see whether our model is acceptable. 

```{r}
table(wdbc.lda.predict.class, wdbc.pcst.test.df$diagnosis)
```

So according to this output, the model predicted 87 times that the diagnosis is 0 (benign) when the actual observation was 0 (benign) and 10 times it predicted incorrectly. Similarly, the model predicted that the diagnosis is 1 (malignant) and once predicted incorrectly. 

The accuracy of this model in predicting benign tumors is 87/97 = 0.8969 or 89.69% accurate.
The accuracy of this model in predicting malignant tumors is 38/39 = 0.9743 or 97.43% accurate.


An advanced way of validating the accuracy of our model is by using a k-fold cross-validation. 

## Conclusion










