---
title: "Logistic Regression"
author: "Shravan Kuchkula"
date: "11/14/2017"
output:
  github_document:
    toc: yes
  html_document:
    keep_md: yes
    theme: cosmo
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
---

## Introduction

Logistic regression, also called a logit model, is used to model dichotomous outcome variables. In the logit model the log odds of the outcome is modeled as a linear combination of the predictor variables.

```{r message=FALSE, warning=FALSE}
# Load all the libraries
installRequiredPackages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[,"Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

libs <- c("dplyr", "tidyr", "ggplot2",
          "magrittr", "markdown", "knitr", "yaml",
           "broom", "aod")

installRequiredPackages(libs)
```

## Problem Statement: 

A researcher is interested in how variables, such as GRE (Graduate Record Exam scores), GPA (grade point average) and prestige of the undergraduate institution, effect admission into graduate school. The response variable, admit/don’t admit, is a binary variable.

## Data description:

Read in the dataset.
```{r}
mydata <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
glimpse(mydata)
```

This dataset has a binary response called `admit`. There are three predictor variables: `gre`, `gpa`, `rank`. The first two are continuous variables. Rank takes the value of 1 through 4. Institutions with a rank 1 have the highest prestige, while those with a rank of 4 have the lowest.

Descriptive statistics, sd and contingency table is shown:

```{r}
summary(mydata)
```

```{r}
sapply(mydata, sd)
```

```{r}
# two way contingency table
xtabs(~admit + rank, data = mydata)
```


## logit method

First lets convert the rank variable to a factor.
```{r}
mydata$rank <- factor(mydata$rank)
```

Next, let's run the glm function
```{r}
mylogit <- glm(admit ~ gre + gpa + rank, data = mydata, family = "binomial")
summary(mylogit)
```







